---
title: "pca_final_project"
author: "Jessica Miller"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
Purpose: Dimensionality reduction with PCA of oceanographic buoy data for use in GLMs.
DataSource: https://www.ndbc.noaa.gov/station_history.php?station=bftn7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Install and load necessary libraries

```{r message=FALSE}
# install.packages("dplyr")
# install.packages("lubridate")
# install.packages("FactoMineR")
# install.packages("factoextra")

library(MASS) # For negative binomial regression
library(dplyr)
library(lubridate)
library(FactoMineR)
library(factoextra)

```

#### Load environmental data

```{r}
buoydata_full <- read.table("bftn7h2010.txt", header=TRUE)

```

#### Select and filter necessary variables
```{r}
buoydata <- buoydata_full %>%
  select(YY, MM, DD, hh, WDIR, WSPD, GST, PRES, ATMP, WTMP) %>%
  filter(WDIR < 999, WSPD < 99, GST < 99, WTMP < 999)
summary(buoydata)
```

#### Inspect variables
```{r}
hist(buoydata$WDIR)
hist(buoydata$WSPD)
hist(buoydata$GST)
hist(buoydata$PRES)
hist(buoydata$ATMP)
hist(buoydata$WTMP)
```

#### Assign Week numbers to dates
```{r}
buoydata <- buoydata %>%
  mutate(Date = make_date(YY, MM, DD)) %>%
   mutate(Week = week(Date))
head(buoydata)
```

#### Average the data by hour and day to reduce variables
```{r}
hourly_avg <- buoydata %>%
  group_by(YY, MM, DD, hh) %>%
  summarise(
    avg_WDIR = mean(WDIR, na.rm = TRUE),
    avg_WSPD = mean(WSPD, na.rm = TRUE),
    avg_GST = mean(GST, na.rm = TRUE),
    avg_PRES = mean(PRES, na.rm = TRUE),
    avg_ATMP = mean(ATMP, na.rm = TRUE),
    avg_WTMP = mean(WTMP, na.rm = TRUE)
  )
head(hourly_avg)
```

#### Remove data columns, select only weeks before and after storms
```{r}
buoydata_before <- buoydata %>%
  filter(Week >= 32 & Week <= 34) %>%
    select(-YY, -MM, -DD, -hh, -Date, -Week) %>%
      mutate(category = "before") %>%
          mutate()


buoydata_after <- buoydata %>%
  filter(Week >= 36 & Week <= 38) %>%
  select(-YY, -MM, -DD, -hh, -Date, -Week) %>%
    mutate(category = "after")

head(buoydata_before)
head(buoydata_after)
```

#### Combine before and after data frames
```{r}
allbuoydata <- rbind(buoydata_before, buoydata_after)
head(allbuoydata)
summary(allbuoydata)
```

#### Run PCA on full dataframe
```{r}
#convert to factor to comply with PCA()
allbuoydata$category <- as.factor(allbuoydata$category)

#select only col 1:6 so PCA ignores category column
pca <- PCA(allbuoydata[,1:6], scale.unit = TRUE, graph = FALSE)
summary(pca)
fviz_pca_biplot(pca, label="var", habillage=allbuoydata$category,
                addEllipses=TRUE, ellipse.level=0.95, col.var="black", 
                palette=c("cadetblue", "darkslateblue"), 
                arrow.size=5, labelsize=5,
                title="Hurricane Earl (2010), PCA Results")
```

#### Run PCA on before and after data separately
```{r}
buoydata_before$category <- as.factor(buoydata_before$category)
pca_before <- PCA(buoydata_before[,1:6], scale.unit = TRUE, graph = FALSE)
summary(pca_before)
fviz_pca_biplot(pca_before, label="var", habillage=buoydata_before$category, 
                col.var="black", palette=c("darkslateblue"),
                arrow.size=5, labelsize=5)

buoydata_after$category <- as.factor(buoydata_after$category)
pca_after <- PCA(buoydata_after[,1:6], scale.unit = TRUE, graph = FALSE)
summary(pca_after)
fviz_pca_biplot(pca_after,label="var", habillage=buoydata_after$category, 
                col.var="black", palette=c("cadetblue"),
                arrow.size=5, labelsize=5)
```

#### Generate Scree Plots
```{r}
# all
eigenvalues <- pca$eig
head(eigenvalues[, 1:2])


barplot(eigenvalues[, 2], names.arg=1:nrow(eigenvalues), 
        main = "Variances - All",
        xlab = "Principal Components",
        ylab = "Percentage of variances",
        col ="steelblue")
lines(x = 1:nrow(eigenvalues), eigenvalues[, 2], 
      type="b", pch=19, col = "deeppink4")

# before

eigenvalues_b <- pca_before$eig
head(eigenvalues_b[, 1:2])


barplot(eigenvalues_b[, 2], names.arg=1:nrow(eigenvalues_b), 
        main = "Variances - Before",
        xlab = "Principal Components",
        ylab = "Percentage of variances",
        col ="steelblue")
# Add connected line segments to the plot
lines(x = 1:nrow(eigenvalues_b), eigenvalues_b[, 2], 
      type="b", pch=19, col = "deeppink4")

# after

eigenvalues_a <- pca_after$eig
head(eigenvalues_a[, 1:2])


barplot(eigenvalues_a[, 2], names.arg=1:nrow(eigenvalues_a), 
        main = "Variances - After",
        xlab = "Principal Components",
        ylab = "Percentage of variances",
        col ="steelblue")
# Add connected line segments to the plot
lines(x = 1:nrow(eigenvalues_a), eigenvalues_a[, 2], 
      type="b", pch=19, col = "deeppink4")

```


```{r}
library(dplyr)

# Assign periods based on weeks
buoydata <- buoydata %>%
  mutate(Period = case_when(
    Week %in% c(32, 33, 34) ~ "Before",
    Week %in% c(36, 37, 38) ~ "After",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(Period)) # Remove rows outside the desired periods

# Perform PCA on buoy data
pca <- prcomp(buoydata %>% select(WDIR, WSPD, GST, PRES, ATMP, WTMP), center = TRUE, scale. = TRUE)
pca_results <- as.data.frame(pca$x)

# Add PCA scores back to the dataset
buoydata <- cbind(buoydata, pca_results)

# Aggregate by Period
buoy_aggregated <- buoydata %>%
  group_by(Period) %>%
  summarise(PC1 = mean(PC1), PC2 = mean(PC2))

print(buoy_aggregated)

#load bridgenet data
bridgenet <- read.csv("bridgenet.csv", header=TRUE)

# Assign periods based on weeks
abundance_data <- bridgenet %>%
  mutate(Period = case_when(
    Week == 28 ~ "Before",
    Week == 49 ~ "After",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(Period)) # Remove rows outside the desired periods

# Aggregate fish data by Period, Taxa, and adjust for volume filtered
abundance_aggregated <- abundance_data %>%
  group_by(Period, Taxa) %>%
  summarise(
    TotalAbundance = sum(TotalAbundance),
    VolumeFiltered = sum(VolumeFiltered)
  )

print(abundance_aggregated)

# Merge the aggregated datasets
combined_data <- abundance_aggregated %>%
  left_join(buoy_aggregated, by = "Period")

print(combined_data)


```

```{r}

# Fit the GLM (using Poisson regression initially)
glm_poisson <- glm(
  TotalAbundance ~ PC1 + PC2 + Period + offset(log(VolumeFiltered)),
  data = combined_data,
  family = poisson()
)

# Check for overdispersion
dispersion <- sum(residuals(glm_poisson, type = "pearson")^2) / df.residual(glm_poisson)
print(dispersion)

# If overdispersion is detected, use Negative Binomial regression
if (dispersion > 1.5) {
  glm_nb <- glm.nb(
    TotalAbundance ~ PC1 + PC2 + Period + offset(log(VolumeFiltered)),
    data = combined_data
  )
  summary(glm_nb)
} else {
  summary(glm_poisson)
}

```

